---
title: "Scaling smotif with smotifrs (Rust backend)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling smotif with smotifrs (Rust backend)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4
)
```

`smotifrs` is the Rust-accelerated companion to [smotif]. It exposes the
same enumeration API for size-2 / size-3 / size-4 motifs but runs the
inner loops on a compact CSR adjacency in Rust, and can read
`nodes.parquet` / `edges.parquet` directly without materializing a
`SpatialGraph` in R memory. This vignette shows how to install both
packages, swap backends with a single argument, read parquet directly
in Rust, and benchmark the speed/memory tradeoff at three dataset
sizes. A short list of natural next steps closes the tutorial.

## 1. Installation

Both packages are needed: `smotif` for the data model and downstream
analyses, `smotifrs` for the Rust kernels.

```r
# 1. Install smotif (R-only, fast)
remotes::install_github("rdries/smotif")

# 2. Install smotifrs — RECOMMENDED: pre-built binary (no Rust toolchain)
install.packages(
  "smotifrs",
  repos = c("https://rdries.r-universe.dev", getOption("repos"))
)

# 2a. Source install (only if you have cargo + rustc on your PATH)
# remotes::install_github("rdries/smotifrs")
```

The R-universe path downloads a pre-compiled `.so` / `.dylib` / `.dll`
matching your platform, so users who don't want to deal with Rust never
have to. Source builds remain available for power users who do.

```{r load, message = FALSE}
library(smotif)
library(smotifrs)
library(ggplot2)
```

## 2. Quickstart: a one-line backend swap

Anywhere you'd call `smotif::find_motifs(sg, ...)`, pass
`backend = "rust"`. The output is a `Motifs` object identical in
schema to the R backend — every downstream consumer (null models,
gene association, niche comparison) works unchanged.

```{r quickstart}
ext <- system.file("extdata", "workshop_xenium", package = "smotif")
sg <- read_spatial_graph(file.path(ext, "nodes.parquet"),
                         file.path(ext, "edges.parquet"))
sg

m_ig <- find_motifs(sg, size = 3L, backend = "igraph")
m_rs <- find_motifs(sg, size = 3L, backend = "rust")

# Bit-for-bit parity on the catalog.
identical(
  m_ig$catalog[order(motif_id), .(motif_id, count)],
  m_rs$catalog[order(motif_id), .(motif_id, count)]
)
```

**What's happening under the hood.** `smotif`'s pure-R backend builds an
igraph object and runs `data.table` joins to enumerate triangles and
wedges. `smotifrs` instead constructs a compact compressed-sparse-row
adjacency in Rust (sorted neighbor lists per vertex), then runs:

* **size 2** — vectorized over edges with packed integer color keys.
* **size 3 closed** — sorted-merge triangle intersection: for every
  edge `u-v` with `u < v`, walk `N(u) ∩ N(v)` keeping only `w > v`.
* **size 3 open** — for each center, every unordered pair of
  neighbors that is *not itself* an edge (one O(1) hash lookup against
  an `ahash` packed-edge set).
* **size 4** — extend every connected 3-subgraph by a fourth vertex,
  dedupe with `u128`-packed keys, classify by `(n_edges, sorted
  degrees)` into one of `{path, claw, cycle, paw, diamond, K4}`.

Canonical motif labels (sort vertices by `(degree-in-motif, color)`)
match the R kernel byte-for-byte, which is why the parity check passes.

## 3. Direct parquet I/O — for million-cell datasets

For graphs that don't fit comfortably in R memory, skip the
`SpatialGraph` round-trip. `find_motifs_from_parquet()` reads
nodes.parquet and edges.parquet directly via `arrow-rs`, builds the
CSR in Rust, and returns the motif catalog without ever allocating
the bulk node/edge data on R's heap.

```{r parquet-direct}
m_direct <- find_motifs_from_parquet(
  file.path(ext, "nodes.parquet"),
  file.path(ext, "edges.parquet"),
  size = 3L
)
identical(
  m_direct$catalog[order(motif_id), count],
  m_rs$catalog[order(motif_id), count]
)
```

The motif catalog and incidence tables are small (one row per motif
type and one per vertex-instance respectively) so they round-trip back
to R cheaply. The expensive 1M+ row edges table stays in Rust.

## 4. Benchmark: three synthetic kNN graphs

We generate three random spatial graphs at increasing sizes and time
both backends at sizes 3 and 4. The R-side memory peak is read from
`gc()`'s `"max used"` column and tells a clean story: the Rust backend
keeps the graph out of R's heap, so it costs near-zero R memory.

```{r helpers}
make_synthetic <- function(n, k = 6L, seed = 1L) {
  set.seed(seed)
  coords <- cbind(x = stats::runif(n, 0, 1000),
                  y = stats::runif(n, 0, 1000))
  types <- sample(c("T", "B", "M", "E", "X"), n, replace = TRUE)
  build_spatial_graph(
    coords, types, "synth",
    cell_id = paste0("c", seq_len(n)),
    method = "knn", k = k
  )
}

peak_mb_from_gc <- function(g) {
  # Robust across R versions: locate the "max used" column and read
  # the Mb counterpart immediately after it.
  i <- which(colnames(g) == "max used")
  if (!length(i) || ncol(g) < i + 1L) return(NA_real_)
  sum(g[, i + 1L])
}

bench_one <- function(label, fn) {
  invisible(gc(reset = TRUE, full = TRUE))
  t1 <- Sys.time()
  # Keep the result alive until t2 so its allocation is captured by gc().
  result_holder <- fn()
  t2 <- Sys.time()
  g <- gc(verbose = FALSE)
  rm(result_holder)
  data.frame(
    label     = label,
    elapsed   = as.numeric(difftime(t2, t1, units = "s")),
    r_peak_mb = peak_mb_from_gc(g)
  )
}
```

```{r benchmarks, results = "hide"}
sizes <- c(small = 2500L, medium = 10000L, large = 30000L)
out <- list()
for (sz_name in names(sizes)) {
  n <- sizes[[sz_name]]
  message("building synthetic graph: ", n, " cells")
  sg_n <- make_synthetic(n)
  for (motif_size in c(3L, 4L)) {
    out[[length(out) + 1L]] <- transform(
      bench_one(
        sprintf("%s/N=%d/size=%d/igraph", sz_name, n, motif_size),
        function() find_motifs(sg_n, size = motif_size, backend = "igraph")
      ),
      backend = "igraph", N = n, motif_size = motif_size, sz_name = sz_name
    )
    out[[length(out) + 1L]] <- transform(
      bench_one(
        sprintf("%s/N=%d/size=%d/rust", sz_name, n, motif_size),
        function() find_motifs(sg_n, size = motif_size, backend = "rust")
      ),
      backend = "rust", N = n, motif_size = motif_size, sz_name = sz_name
    )
  }
}
bench <- do.call(rbind, out)
bench$sz_name <- factor(bench$sz_name, levels = names(sizes))
```

```{r bench-table}
data.table::setDT(bench)
bench[, .(label, elapsed_s = round(elapsed, 2),
          r_peak_mb = round(r_peak_mb, 1))]
```

### 4a. Speedup vs dataset size

For each motif size, the Rust backend's win grows with N because the R
backend allocates increasingly large intermediate `data.table`s during
joins, while the Rust CSR + ahash lookups stay constant-factor.

```{r time-plot, fig.width = 7, fig.height = 4.2}
ggplot(bench,
       aes(x = N, y = elapsed, color = backend, linetype = backend,
           shape = backend)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.4) +
  facet_wrap(~ paste("size", motif_size), scales = "free_y") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c(igraph = "#888888", rust = "#d73027")) +
  labs(x = "number of cells", y = "elapsed time (s, log scale)",
       title = "Wall-clock: pure-R igraph vs Rust backend") +
  theme_minimal(base_size = 10)
```

```{r speedup-table}
sp <- data.table::dcast(bench,
                        sz_name + N + motif_size ~ backend,
                        value.var = "elapsed")
sp[, speedup := round(igraph / rust, 2)]
sp[, .(sz_name, N, motif_size,
       igraph_s = round(igraph, 2),
       rust_s   = round(rust, 2),
       speedup)]
```

### 4b. R-side memory peak

Pure R routes pass through `data.table` joins that materialize every
candidate motif as a row before deduplication; the Rust backend keeps
the graph in its own heap and only ships the final catalog/incidence
back to R. The result is a roughly N-proportional gap in R-side
working set.

```{r mem-plot, fig.width = 7, fig.height = 4.2}
ggplot(bench,
       aes(x = sz_name, y = r_peak_mb, fill = backend)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ paste("size", motif_size)) +
  scale_fill_manual(values = c(igraph = "#888888", rust = "#d73027")) +
  labs(x = "dataset size", y = "R-heap peak (MiB)",
       title = "R-side memory peak (gc max-used)",
       subtitle = "Rust kernel keeps the graph out of R memory") +
  theme_minimal(base_size = 10)
```

The reported number is the maximum R heap used during the call (read
from `gc()`'s `"max used"` column). It excludes whatever Rust's own
allocator reserves for its CSR — that's a separate working set, far
smaller than the equivalent `data.table` representation in R, but not
visible to R's `gc()`.

## 5. Where this matters in practice

* **Hot inner loop in repeated analyses.** Permutation-based
  enrichment (`smotif::test_motif_enrichment`) doesn't currently
  re-enumerate motifs (it just relabels), but if you script your own
  motif-recount-per-permutation, swapping the backend cuts wall time.
* **Parquet pipelines from upstream tools.** If you're already writing
  `nodes.parquet` and `edges.parquet` from a Python or Giotto pipeline,
  `find_motifs_from_parquet()` lets you skip the R-side ingest
  entirely. Useful when those files exist on object storage and you'd
  rather not pull them through R's Arrow bindings.
* **CI / batch jobs at scale.** The 30k benchmark runs in seconds with
  the Rust backend on a single core; parallelism (next section) makes
  ≥1M-cell jobs feasible inside R.

## 6. Future improvements

A few natural next steps, in rough order of payoff for the size of
slides we expect to see:

1. **Parallel enumeration** — the current Rust backend is
   single-threaded. The size-3-closed and size-3-open kernels are
   embarrassingly parallel over the outer vertex loop; wiring up
   `rayon::into_par_iter` is a one-liner per kernel and should
   give 4–8× on a typical laptop CPU.
2. **Streaming size-4 incidence to disk.** Catalogs stay small even
   at 1M cells, but the per-instance incidence table can balloon into
   the hundreds of millions of rows. An `incidence_path = ...`
   argument that streams parquet output (one batch per topology
   class) keeps R's memory flat and makes downstream queries a
   `arrow::open_dataset()` call.
3. **Size 5+ motifs via ESU + color-coding.** The exhaustive
   3-subgraph extension we use for size 4 doesn't generalize cleanly;
   Wernicke's ESU enumerator with color-coding samples a controlled
   fraction of the search space and is the standard approach for k≥5.
4. **Per-sample meta-analysis in Rust.** `associate_genes(pooling = "per_sample")`
   loops over samples in R today; pushing the Wilcoxon rank-sum down
   to Rust (with sample-stratified rank arrays) avoids the
   per-call extendr overhead.
5. **Exact canonical labeling for size 4.** Today's `(degree, color)`
   sort is approximate (collides on certain colored cycles). Bliss or
   a hand-coded Schreier-Sims tailored to size 4 would close the gap
   without the cost of a full nauty-style canonicalization.
6. **GPU enumeration.** For very large slides, a CUDA / wgpu kernel
   for triangle counting + per-pair color aggregation would push
   throughput by another order of magnitude — useful once whole-organ
   atlases (millions of cells per slide, dozens of slides) become
   routine.

The first two are the most concrete wins; everything below is
research-tier work whose payoff depends on what biology people
actually want to ask of these datasets.
